\texttt{Davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-30 paper/main.tex
cleaned up a bunch of LaTeX stuff

When accessed (unless its value has been set to \texttt{None}), \texttt{davos.project} will evaluate to a \texttt{Project} object that represents the project used by the current notebook (strings assigned to \texttt{davos.project} are converted to \texttt{Project}s internally). This object supports methods for interacting with the current project, including locating its directory within the file system, listing all installed packages' names and versions, changing the project's name, and deleting its contents.
\texttt{Project} instances can also be created and managed programmatically, and \texttt{Davos} provides additional utilities for viewing and working with all existing projects (see Secs.~\ref{subsec:config} and \ref{subsec:toplevel}).

@@

\item \texttt{.project}: This attribute's value is a \texttt{Project} instance representing the \texttt{Davos} project associated with the current notebook.
  As described in Section~\ref{subsec:projects}, \texttt{Davos} projects serve to isolate packages installed by \texttt{smuggle} statements from the user's main Python environment, and the \texttt{Project} class provides an interface for inspecting and managing projects at runtime.
  This attribute's default value is a notebook-specific project named for the absolute path to the notebook file.
  To change the project used in the current notebook (e.g., in order to use the same project in multiple related notebooks), this attribute may be assigned a different \texttt{Project} instance or, for simplicity, the name of the desired project as a string or \texttt{pathlib.Path} (either of which will be converted to a \texttt{Project} on assignment).
  Alternatively, setting \texttt{davos.project} to \texttt{None} will disable project-based isolation for the current notebook and cause \texttt{Davos} to install any missing packages directly into the main Python environment.
  This attribute can be reset to its default value using the top-level \texttt{use\_default\_project()} function (see Sec.~\ref{subsec:toplevel}).
  For more information about \texttt{Davos} projects, see Section \ref{subsec:projects}.

@@

\item \texttt{.environment}: This attribute's value is a string denoting the set of environment-dependent ``helper functions'' used by \texttt{Davos} in the current notebook.
  As described in Section \ref{sec:architecture}, \texttt{Davos} internally chooses between interchangeable implementations of certain core features based on various properties of the notebook's frontend and IPython kernel.
  As of this writing, three unique combinations of helper functions are required to support existing notebook environments, ergo this attribute has three possible values: \texttt{"IPython<7.0"}, \texttt{"IPython>=7.0"}, or \texttt{"Colaboratory"}.
  However, this attribute could take on additional values in the future as new notebook interfaces are created and IPython's internals are updated, and as additional versions of helper functions are added to \texttt{Davos} to support them.

@@

\item \texttt{configure(**kwargs)}: This function provides an alternate way of assigning values to the writeable attributes listed in Section \ref{subsec:config} and can be used to configure multiple options at once (see Sec.~\ref{sec:illustrative-example} for example usage).
  The function accepts attribute names as keyword-only arguments to which their desired values are passed.
  If any of the options passed are incompatible (e.g., both \texttt{confirm\_install=True} and \texttt{noninteractive=True} are passed) or assignment to any of the specified attributes fails for any reason, none of the given options will be modified.

@@

\item \texttt{require\_python(version\_spec, warn=False, extra\_msg=None, pre\-re\-leases=\\None)}: Through \texttt{smuggle} statements and onion comments, \texttt{Davos} can automatically ensure that all Python packages needed to run a notebook are installed, and that the same versions of those packages are used no matter when or by whom the notebook is run.
  However, because \texttt{Davos} operates at runtime, one thing it cannot do automatically is install and switch to a specific version of Python itself.
  Distributing shared code along with a precise Python version for running it requires a heavier-weight solution, such as a Conda environment or Docker container (see Fig.~\ref{fig:code-sharing}).
  Yet a \texttt{Davos}-enhanced notebook may still \texttt{smuggle} certain packages that depend on users having a particular Python version or range of versions (e.g., even just within the standard library, the \texttt{dataclass} module was first added in Python 3.7 \cite{Smit17} and at least 19 modules are slated for removal in Python 3.13 \cite{HeimCann19}).
  The \texttt{davos.require\_python()} function can be added to the top of a \texttt{Davos}-enhanced notebook to communicate to users that the notebook's code should be run with a specific or constrained Python version (see Sec.~\ref{sec:illustrative-example} for example usage).
  The function may be passed a version identifier (e.g., \texttt{"3.10.5"}) or any valid version specifier \cite{CoghStuf13} (e.g., \texttt{"\raisebox{0.5ex}{\texttildelow}=3.11"}, \texttt{">=3.9;<3.12"}, etc.) and will raise an exception if the user's Python version is incompatible.
  Alternatively, a ``soft'' or suggested constraint can be imposed by passing \texttt{warn=True} to issue a warning rather than raise an error.
  Additional information can be added to the default error/warning message (e.g., the specific reason for this requirement) via the \texttt{extra\_msg} argument, and the optional \texttt{prereleases} argument can be used to explicitly allow (\texttt{True}) or disallow (\texttt{False}) pre-release versions (by default, the policy is determined by the value of \texttt{version\_spec}).

\item \texttt{use\_default\_project()}: By default, each \texttt{Davos}-enhanced notebook will create and use a notebook-specific project named based on its absolute path.
  If a user manually changes the project used by the current notebook (i.e., by setting the value of the \texttt{davos.project} attribute; see Sec.~\ref{subsec:config}), this function can be called to switch back to using the notebook's default project and reset \texttt{davos.project} to its default value.
  See Section \ref{subsec:projects} for more information about \texttt{Davos} projects and Section \ref{sec:illustrative-example} for an illustrative example.

@@

The example code throughout Section \ref{subsec:onion} illustrates a typical use case that we envision for \texttt{Davos}: a series of smuggle statements and onion comments with version specifiers or other options collectively describes and automatically constructs a reproducible environment for running the code that follows it.
When added to the top of a Jupyter notebook, this allows researchers to bundle their code and its dependencies into a single file that can be easily shared and run without any additional tools or setup, automatically installs its required packages at runtime, isolates them from the user's main Python environment, and ensures their versions do not change unexpectedly over time.
In this section, we have contrived a more complex scenario to highlight some of \texttt{Davos}'s more advanced features, and illustrate how they may be used to handle certain challenges that can arise when writing, running, and sharing reproducible scientific code.

@@

After installing and importing \texttt{Davos} (lines 1--2), we first use the \texttt{davos.require\_\-python()} function to constrain the Python version used to run the notebook (see Sec.~\ref{subsec:toplevel}).
As described above, the example code in Figure \ref{fig:illustrative-example} loads two different versions of the \texttt{pandas} library: first, an older version needed to access a dataset saved in an outmoded format, then a newer one to use throughout the remainder of the notebook.
We therefore want to make sure upfront (in line 6) that the notebook's Python version falls within the range of versions that both of these two versions of \texttt{pandas} support.
If it does not, the function in line 6 will raise an error that includes a message to this effect (lines 4--5).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}

@@

Lines 30--37 account for both of these possibilities programmatically:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
First, in line 30, we set the \texttt{davos.project} attribute to \texttt{None} to temporarily
allow installing smuggled packages outside of the notebook's project directory.
As noted in Section \ref{subsec:projects}, this is typically discouraged, as doing so can risk interfering with the user's Python environment if existing package versions are overwritten.
In this particular case, however, a combination of factors make this relatively safe and inconsequential.
First, the package we need to install directly into the notebook server environment (\texttt{widgetsnbextension}) is smuggled without an accompanying onion comment (line 34), meaning that \texttt{Davos} will not replace any version the user may already have installed.
Second, the package has no dependencies of its own, so if \texttt{Davos} does install it, no other packages will be installed or updated as a side effect.
Third, the package itself provides no functionality outside of rendering Jupyter widgets, so its presence would not alter any other code's expected behavior.

@@

Next, we load in the reformatted dataset (line 41) and pre-trained
model (line 43) that we wish to use in our analysis. In our
hypothetical example, we can suppose that the model was provided as a
\texttt{scikit-learn} \texttt{Pipeline} object that passes data
through two pre-trained models in succession. First, a trained \texttt{CountVectorizer}
instance converts text data to an array of word counts.  The
word counts are then passed to a topic model~\cite{BleiEtal03} using a
pre-trained \texttt{LatentDirichletAllocation} instance.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example8}
\end{center}
Let us suppose that the \texttt{Pipeline} object had been saved by its
original creator using the \texttt{joblib} package, as
\texttt{scikit-learn}'s documentation recommends \cite{skle22}. Because
\texttt{joblib} uses the \texttt{pickle} protocol internally, the
ability to save and load pre-trained models is not guaranteed across
different \texttt{scikit-learn} versions. For example, suppose that
the \texttt{Pipeline} object was created using \texttt{scikit-learn}
v0.21.3. In that version (and previous versions) of \texttt{scikit-learn}, the
\texttt{LatentDirichletAllocation} class was defined in
the \texttt{sklearn.de\-comp\-o\-si\-tion.online\_lda} module. However, in version
0.22.0, that module was renamed to ``\texttt{\_online\_lda}'' and in
version 0.22.1, it was again renamed to ``\texttt{\_lda}.''

@@

First, \texttt{smuggle} statements and onion comments enable researchers to specify their notebooks' dependencies directly within the code that requires them (see Secs.~\ref{subsec:smuggle} and  \ref{subsec:onion}).
This eliminates the need to either manually create and maintain separate configuration files or use an additional tool to generate them, and to then distribute these files alongside their primary code base for users to download.

@@

While \texttt{Davos} enables developers to conveniently specify all project
dependencies, there are some edge cases and limitations that are worth
considering.
First, prior studies on reproducibility of Jupyter notebooks (e.g., \cite{PimeEtal19}) identified a key challenge in the fact that, unlike Python scripts, notebook cells may be manually executed in a non-linear order, which may differ from the order in which they were executed by the notebook's original author.
This can result in various ``hidden state'' problems, such as variables referencing different objects than the currently running cell expects them to, given its point in the notebook's execution order.
In theory, using \texttt{Davos} to \texttt{smuggle} different versions of the same package across different notebook cells could exacerbate this issue if a user executed those cells out of their intended order.
Therefore an important consideration when using \texttt{Davos} is that it is perhaps even more important to execute notebook cells in order than would be the case in the standard (non-\texttt{Davos}) setup.
However, a few ``best practices'' may be adopted to help mitigate this risk.
First, when sharing a \texttt{Davos}-enhanced notebook in a state that is intended to be reproducible, code cells should appear in their intended execution order, such that users can successfully run them top to bottom.
In a scenario where this is impossible, the intended order should be clearly indicated in code comments and/or markdown cells.
This is already a common practice for notebooks that include system calls to install required packages and dependencies, and the same approach would work well in \texttt{Davos}-enhanced notebooks.
Second, while \texttt{Davos} makes it possible to \texttt{smuggle} multiple versions of a package within the same notebook, users should prefer to split code that requires different versions of the same package into different notebooks.
In addition to mitigating the issue described above, this will also allow \texttt{Davos} to store a single intended version of the package in each notebook's project directory, rather than reinstalling the package each time a different version is smuggled.
Finally, in situations where it \textit{is} necessary or preferable to temporarily \texttt{smuggle} a different version of a particular package than is used throughout the rest of the notebook (e.g., Fig.~\ref{fig:illustrative-example}), the code to \texttt{smuggle} that temporary version, perform all necessary operations, and (re-)\texttt{smuggle} the ``default'' version should be contained within a single notebook cell, such that (absent any errors in its execution) the same version is always installed and imported before and after the cell is run.
For example, in Figure~\ref{fig:illustrative-example}, lines 14--19 should be run within a single cell, and lines 42--44 should be run in a single cell.

A second limitation of \texttt{Davos} relates to how packages are installed and managed.
As of this writing, \texttt{Davos} can install packages using \texttt{pip}, but not
other standard Python package management systems such as \texttt{conda}.
Therefore packages that are not installable via \texttt{pip} are currently
unsupported by \texttt{Davos}. We anticipate adding support for other package management
systems, including \texttt{conda} \cite{Anac12}, in a future release.

@@

Our work was supported in part by NSF grant number 2145172 to JRM.
The content is solely the responsibility of the authors and does not
necessarily represent the official views of our supporting
organizations.

>>>>

When accessed (unless its value has been set to \texttt{None}), \texttt{davos.project} will evaluate to a \texttt{Project} object that represents the project used by the current notebook (strings assigned to \texttt{davos.project} are converted to \texttt{Project}s internally). This object supports methods for interacting with the current project, including locating its directory within the file system, listing all installed packages' names and versions, changing the project's name, and deleting its contents.
\texttt{Project} instances can also be created and managed programmatically, and \texttt{Davos} provides additional utilities for viewing and working with all existing projects (see Secs.~\ref{subsec:config} and~\ref{subsec:toplevel}).

@@

\item \texttt{.project}: This attribute's value is a \texttt{Project} instance representing the \texttt{Davos} project associated with the current notebook.
  As described in Section~\ref{subsec:projects}, \texttt{Davos} projects serve to isolate packages installed by \texttt{smuggle} statements from the user's main Python environment, and the \texttt{Project} class provides an interface for inspecting and managing projects at runtime.
  This attribute's default value is a notebook-specific project named for the absolute path to the notebook file.
  To change the project used in the current notebook (e.g., in order to use the same project in multiple related notebooks), this attribute may be assigned a different \texttt{Project} instance or, for simplicity, the name of the desired project as a string or \texttt{pathlib.Path} (either of which will be converted to a \texttt{Project} on assignment).
  Alternatively, setting \texttt{davos.project} to \texttt{None} will disable project-based isolation for the current notebook and cause \texttt{Davos} to install any missing packages directly into the main Python environment.
  This attribute can be reset to its default value using the top-level \texttt{use\_default\_project()} function (see Sec.~\ref{subsec:toplevel}).
  For more information about \texttt{Davos} projects, see Section~\ref{subsec:projects}.

@@

\item \texttt{.environment}: This attribute's value is a string denoting the set of environment-dependent ``helper functions'' used by \texttt{Davos} in the current notebook.
  As described in Section~\ref{sec:architecture}, \texttt{Davos} internally chooses between interchangeable implementations of certain core features based on various properties of the notebook's frontend and IPython kernel.
  As of this writing, three unique combinations of helper functions are required to support existing notebook environments, ergo this attribute has three possible values: \texttt{"IPython<7.0"}, \texttt{"IPython>=7.0"}, or \texttt{"Colaboratory"}.
  However, this attribute could take on additional values in the future as new notebook interfaces are created and IPython's internals are updated, and as additional versions of helper functions are added to \texttt{Davos} to support them.

@@

\item \texttt{configure(**kwargs)}: This function provides an alternate way of assigning values to the writeable attributes listed in Section~\ref{subsec:config} and can be used to configure multiple options at once (see Sec.~\ref{sec:illustrative-example} for example usage).
  The function accepts attribute names as keyword-only arguments to which their desired values are passed.
  If any of the options passed are incompatible (e.g., both \texttt{confirm\_install=True} and \texttt{noninteractive=True} are passed) or assignment to any of the specified attributes fails for any reason, none of the given options will be modified.

@@

\item \texttt{require\_python(version\_spec, warn=False, extra\_msg=None, pre\-re\-leases=\\None)}: Through \texttt{smuggle} statements and onion comments, \texttt{Davos} can automatically ensure that all Python packages needed to run a notebook are installed, and that the same versions of those packages are used no matter when or by whom the notebook is run.
  However, because \texttt{Davos} operates at runtime, one thing it cannot do automatically is install and switch to a specific version of Python itself.
  Distributing shared code along with a precise Python version for running it requires a heavier-weight solution, such as a Conda environment or Docker container (see Fig.~\ref{fig:code-sharing}).
  Yet a \texttt{Davos}-enhanced notebook may still \texttt{smuggle} certain packages that depend on users having a particular Python version or range of versions (e.g., even just within the standard library, the \texttt{dataclass} module was first added in Python 3.7~\cite{Smit17} and at least 19 modules are slated for removal in Python 3.13~\cite{HeimCann19}).
  The \texttt{davos.require\_python()} function can be added to the top of a \texttt{Davos}-enhanced notebook to communicate to users that the notebook's code should be run with a specific or constrained Python version (see Sec.~\ref{sec:illustrative-example} for example usage).
  The function may be passed a version identifier (e.g., \texttt{"3.10.5"}) or any valid version specifier~\cite{CoghStuf13} (e.g., \texttt{"\raisebox{0.5ex}{\texttildelow}=3.11"}, \texttt{">=3.9;<3.12"}, etc.) and will raise an exception if the user's Python version is incompatible.
  Alternatively, a ``soft'' or suggested constraint can be imposed by passing \texttt{warn=True} to issue a warning rather than raise an error.
  Additional information can be added to the default error/warning message (e.g., the specific reason for this requirement) via the \texttt{extra\_msg} argument, and the optional \texttt{prereleases} argument can be used to explicitly allow (\texttt{True}) or disallow (\texttt{False}) pre-release versions (by default, the policy is determined by the value of \texttt{version\_spec}).

\item \texttt{use\_default\_project()}: By default, each \texttt{Davos}-enhanced notebook will create and use a notebook-specific project named based on its absolute path.
  If a user manually changes the project used by the current notebook (i.e., by setting the value of the \texttt{davos.project} attribute; see Sec.~\ref{subsec:config}), this function can be called to switch back to using the notebook's default project and reset \texttt{davos.project} to its default value.
  See Section~\ref{subsec:projects} for more information about \texttt{Davos} projects and Section~\ref{sec:illustrative-example} for an illustrative example.

@@

The example code throughout Section~\ref{subsec:onion} illustrates a typical use case that we envision for \texttt{Davos}: a series of smuggle statements and onion comments with version specifiers or other options collectively describes and automatically constructs a reproducible environment for running the code that follows it.
When added to the top of a Jupyter notebook, this allows researchers to bundle their code and its dependencies into a single file that can be easily shared and run without any additional tools or setup, automatically installs its required packages at runtime, isolates them from the user's main Python environment, and ensures their versions do not change unexpectedly over time.
In this section, we have contrived a more complex scenario to highlight some of \texttt{Davos}'s more advanced features, and illustrate how they may be used to handle certain challenges that can arise when writing, running, and sharing reproducible scientific code.

@@

After installing and importing \texttt{Davos} (lines 1--2), we first use the \texttt{davos.require\_\-python()} function to constrain the Python version used to run the notebook (see Sec.~\ref{subsec:toplevel}).
As described above, the example code in Figure~\ref{fig:illustrative-example} loads two different versions of the \texttt{pandas} library: first, an older version needed to access a dataset saved in an outmoded format, then a newer one to use throughout the remainder of the notebook.
We therefore want to make sure upfront (in line 6) that the notebook's Python version falls within the range of versions that both of these two versions of \texttt{pandas} support.
If it does not, the function in line 6 will raise an error that includes a message to this effect (lines 4--5).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}

@@

Lines 30--37 account for both of these possibilities programmatically:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
First, in line 30, we set the \texttt{davos.project} attribute to \texttt{None} to temporarily
allow installing smuggled packages outside of the notebook's project directory.
As noted in Section~\ref{subsec:projects}, this is typically discouraged, as doing so can risk interfering with the user's Python environment if existing package versions are overwritten.
In this particular case, however, a combination of factors make this relatively safe and inconsequential.
First, the package we need to install directly into the notebook server environment (\texttt{widgetsnbextension}) is smuggled without an accompanying onion comment (line 34), meaning that \texttt{Davos} will not replace any version the user may already have installed.
Second, the package has no dependencies of its own, so if \texttt{Davos} does install it, no other packages will be installed or updated as a side effect.
Third, the package itself provides no functionality outside of rendering Jupyter widgets, so its presence would not alter any other code's expected behavior.

@@

Next, we load in the reformatted dataset (line 41) and pre-trained
model (line 43) that we wish to use in our analysis. In our
hypothetical example, we can suppose that the model was provided as a
\texttt{scikit-learn} \texttt{Pipeline} object that passes data
through two pre-trained models in succession. First, a trained \texttt{CountVectorizer}
instance converts text data to an array of word counts. The
word counts are then passed to a topic model~\cite{BleiEtal03} using a
pre-trained \texttt{LatentDirichletAllocation} instance.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example8}
\end{center}
Let us suppose that the \texttt{Pipeline} object had been saved by its
original creator using the \texttt{joblib} package, as
\texttt{scikit-learn}'s documentation recommends~\cite{skle22}. Because
\texttt{joblib} uses the \texttt{pickle} protocol internally, the
ability to save and load pre-trained models is not guaranteed across
different \texttt{scikit-learn} versions. For example, suppose that
the \texttt{Pipeline} object was created using \texttt{scikit-learn}
v0.21.3. In that version (and previous versions) of \texttt{scikit-learn}, the
\texttt{LatentDirichletAllocation} class was defined in
the \texttt{sklearn.de\-comp\-o\-si\-tion.online\_lda} module. However, in version
0.22.0, that module was renamed to ``\texttt{\_online\_lda}'' and in
version 0.22.1, it was again renamed to ``\texttt{\_lda}.''

@@

First, \texttt{smuggle} statements and onion comments enable researchers to specify their notebooks' dependencies directly within the code that requires them (see Secs.~\ref{subsec:smuggle} and~\ref{subsec:onion}).
This eliminates the need to either manually create and maintain separate configuration files or use an additional tool to generate them, and to then distribute these files alongside their primary code base for users to download.

@@

While \texttt{Davos} enables developers to conveniently specify all project
dependencies, there are some edge cases and limitations that are worth
considering.
First, prior studies on reproducibility of Jupyter notebooks~\cite[e.g.,][]{PimeEtal19}) identified a key challenge in the fact that, unlike Python scripts, notebook cells may be manually executed in a non-linear order, which may differ from the order in which they were executed by the notebook's original author.
This can result in various ``hidden state'' problems, such as variables referencing different objects than the currently running cell expects them to, given its point in the notebook's execution order.
In theory, using \texttt{Davos} to \texttt{smuggle} different versions of the same package across different notebook cells could exacerbate this issue if a user executed those cells out of their intended order.
Therefore an important consideration when using \texttt{Davos} is that it is perhaps even more important to execute notebook cells in order than would be the case in the standard (non-\texttt{Davos}) setup.
However, a few ``best practices'' may be adopted to help mitigate this risk.
First, when sharing a \texttt{Davos}-enhanced notebook in a state that is intended to be reproducible, code cells should appear in their intended execution order, such that users can successfully run them top to bottom.
In a scenario where this is impossible, the intended order should be clearly indicated in code comments and/or markdown cells.
This is already a common practice for notebooks that include system calls to install required packages and dependencies, and the same approach would work well in \texttt{Davos}-enhanced notebooks.
Second, while \texttt{Davos} makes it possible to \texttt{smuggle} multiple versions of a package within the same notebook, users should prefer to split code that requires different versions of the same package into different notebooks.
In addition to mitigating the issue described above, this will also allow \texttt{Davos} to store a single intended version of the package in each notebook's project directory, rather than reinstalling the package each time a different version is smuggled.
Finally, in situations where it \textit{is} necessary or preferable to temporarily \texttt{smuggle} a different version of a particular package than is used throughout the rest of the notebook (e.g., Fig.~\ref{fig:illustrative-example}), the code to \texttt{smuggle} that temporary version, perform all necessary operations, and (re-)\texttt{smuggle} the ``default'' version should be contained within a single notebook cell, such that (absent any errors in its execution) the same version is always installed and imported before and after the cell is run.
For example, in Figure~\ref{fig:illustrative-example}, lines 14--19 should be run within a single cell, and lines 42--44 should be run in a single cell.

A second limitation of \texttt{Davos} relates to how packages are installed and managed.
As of this writing, \texttt{Davos} can install packages using \texttt{pip}, but not
other standard Python package management systems such as \texttt{conda}.
Therefore packages that are not installable via \texttt{pip} are currently
unsupported by \texttt{Davos}. We anticipate adding support for other package management
systems, including \texttt{conda}~\cite{Anac12}, in a future release.

@@

Our work was supported in part by NSF grant number 2145172 to JRM\@.
The content is solely the responsibility of the authors and does not
necessarily represent the official views of our supporting
organizations.
